Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-07-21 text-backup.tex
In process- editing Andrews Comments

>>>>

%% COASTAL MARGIN FROM INTRO:

Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean's surface but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Data from remote sensing suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023}. Climate change is predicted to have a lasting impact on phytoplankton communities, including shifts in community composition and diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.

, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}

The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated. This underwater topography, in combination with high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}. % Describe general seasonal cycle with functional groups etc.
